Face Recognition: Too Bias, or Not Too Bias?

### 2020-04-20 cvprworkshop.tex

% \author{First Author\\
% Institution1\\
% Institution1 address\\
% {\tt\small firstauthor@i1.org}
% % For a paper whose authors are all at the same institution,
% % omit the following lines up until the closing ``}''.
% % Additional authors and addresses can be added with ``\and'',
% % just like the second author.
% % To save space, use either the email address or home page, not both
% \and
% Second Author\\
% Institution2\\
% First line of institution2 address\\
% {\tt\small secondauthor@i2.org}
% }

\maketitle
%\thispagestyle{empty}

%%%%%%%%% ABSTRACT
\begin{abstract}
We reveal critical insights into problems of bias in state-of-the-art \gls{fr} systems using a novel \gls{bfw} dataset: data balanced for gender and ethnic groups. We show variations in the optimal scoring threshold for face-pairs across different subgroups. Thus, the conventional approach of learning a  global threshold for all pairs resulting in performance gaps among subgroups. By learning subgroup-specific thresholds, we not only mitigate problems in performance gaps but also show a notable boost in the overall performance. Furthermore, we do a human evaluation to measure the bias in humans, which supports the hypothesis that such a bias exists in human perception. For the \gls{bfw} database, source code, and more, visit \href{https://github.com/visionjo/facerec-bias-bfw}{https://github.com/visionjo/facerec-bias-bfw}.
% \href{https://github.com/visionjo/facerec-bias-bfw}{github.com/visionjo/facerec-bias-bfw}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{bibunit}
\input{sections/introduction}
\input{sections/relatedworks}

@@

\begin{figure}[t!]
       \centering
    \includegraphics[width=.95\linewidth]{figures/adaptive_fpr_percent_diff_complete.pdf}
    \caption{\small{\textbf{Percent difference from intended \gls{fpr}.} \emph{Top:} $t_g$ yields \gls{fpr} the span as large as 2$\times$ (\ie 200\%) that intended (\ie \gls{wm} for 1e-4). Furthermore, \gls{f} subgroups tend to perform worse than intended for all cases (while \gls{m} tend to overshoot intended performance, with exception of \gls{im} in for \gls{fpr}=1e-4). \emph{Bottom:} Subgroup-specific thresholds reduces this difference to near zero, where there are small differences, the percent difference across different subgroups is fair (\ie FPR=1e-4).}}\label{fig:percent:difference}
\end{figure}

@@

\glsresetall
\section{Conclusion}
We introduce a new data set \gls{bfw} with eight subgroups balanced across gender and ethnicity. With this, and upon highlighting the challenges and shortcomings of grouping subjects as a single subset, we provide evidence that forming subgroups is meaningful, as the FR algorithm rarely makes mistakes across subgroups. We used an \textit{off-the-shelf} Sphereface, hypothesizing this SOTA CNN suffers from bias because of the imbalanced train-set. Once established that the results do suffer from problems of bias, we observed that the same threshold across ethnic and gender subgroups leads to differences in the \gls{fpr} up to a factor of two. Furthermore, the percent difference from in which is seemingly the cause of the frenzy about bias in FR in main-stream media. Furthermore, we ameliorate these differences with a per-subgroup threshold, leveling out FPR, and achieving a higher TPR. We hypothesized that most humans grew amongst more than their own demographic and, therefore, effectively learn from imbalanced datasets-- a human evaluation validated that humans are biased, as most recognized their personal demographic best. The focused research findings presented here, along with the public database included, are extendable in vast ways. Thus, we see this as the slither to a much larger problem of bias in ML.

{\small
\bibliographystyle{ieee_fullname}
\bibliography{references}
}

>>>>

\maketitle

%%%%%%%%% ABSTRACT
\begin{abstract}
We reveal critical insights into problems of bias in state-of-the-art \gls{fr} systems using a novel \gls{bfw} dataset: data balanced for gender and ethnic groups. We show variations in the optimal scoring threshold for face-pairs across different subgroups. Thus, the conventional approach of learning a  global threshold for all pairs results in performance gaps between subgroups. By learning subgroup-specific thresholds, we reduce performance gaps, and also show a notable boost in overall performance. Furthermore, we do a human evaluation to measure bias in humans, which supports the hypothesis that an analogous bias exists in human perception. For the \gls{bfw} database, source code, and more, visit \href{https://github.com/visionjo/facerec-bias-bfw}{https://github.com/visionjo/facerec-bias-bfw}.
% \href{https://github.com/visionjo/facerec-bias-bfw}{github.com/visionjo/facerec-bias-bfw}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{sections/introduction}
\input{sections/relatedworks}

@@

\begin{figure}[t!]
       \centering
    \includegraphics[width=.95\linewidth]{figures/fpr_percent_diff-crop.pdf}
    \caption{\small{\textbf{Percent difference from intended \gls{fpr}.} \emph{Top:} $t_g$ yields \gls{fpr} the span as large as 2$\times$ (\ie 200\%) that intended (\ie \gls{wm} for 1e-4). Furthermore, \gls{f} subgroups tend to perform worse than intended for all cases (while \gls{m} tend to overshoot intended performance, with exception of \gls{im} in for \gls{fpr}=1e-4). \emph{Bottom:} Subgroup-specific thresholds reduces this difference to near zero, where there are small differences, the percent difference across different subgroups is fair (\ie FPR=1e-4).}}\label{fig:percent:difference}
\end{figure}

@@

\glsresetall
\section{Conclusion}
We introduce a new data set \gls{bfw} with eight subgroups balanced across gender and ethnicity. With this, and upon highlighting the challenges and shortcomings of grouping subjects as a single subset, we provide evidence that forming subgroups is meaningful, as the FR algorithm rarely makes mistakes across subgroups. We used an \textit{off-the-shelf} Sphereface, hypothesizing this SOTA CNN suffers from bias because of the imbalanced train-set. Once established that the results do suffer from problems of bias, we observed that the same threshold across ethnic and gender subgroups leads to differences in the \gls{fpr} up to a factor of two. Also, the percent difference that seemingly the cause of the frenzy about bias in FR in main-stream media. Furthermore, we ameliorate these differences with a per-subgroup threshold, leveling out \gls{fpr}, and achieving a higher \gls{tpr}. We hypothesized that most humans grew among more than their own demographic and, therefore, effectively learn from imbalanced datasets-- a human evaluation validated that humans are biased, as most recognized their personal demographic best. The focused research findings presented here, along with the public database included, are extendable in vast ways. Thus, we see this as the slither to a much larger problem of bias in ML.

{\scriptsize
\bibliographystyle{ieee_fullname}
\bibliography{references}
}

### 2020-04-20 sections/relatedworks.tex
cleanup; removed -vspaces

\section{Background Information}
    \subsection{Bias in \gls{ml}}
        The progress and the commercial value of \gls{ml} is exciting; however, the trust of society in \gls{ml} will take time to build due to inherent biases. The exact definitions and implications of bias vary between sources, as do its sources and types. A common theme is that bias hinders performance ratings in ways that skew to a particular sub-population-- the source varies, whether from humans~\cite{windmann1998subconscious}, data and label types~\cite{tommasi2017deeper}, \gls{ml} models~\cite{amini2019uncovering, kim2019learning}, or evaluation protocols~\cite{stock2018convnets}. For instance, a vehicle-detection model might miss cars if training data was mostly trucks. In practice, many \gls{ml} systems learn on biased data, which could be detrimental for society.

    \subsection{Bias in \gls{fr}}
        Problems of bias in \gls{fr} have been driven by different motivations; therefore, solutions have been posed for different matters. To name a few: data augmentation~\cite{yin2019feature}, one-shot learning~\cite{ding2018one}, demographic parity and fairness with priority on privacy~\cite{huang2018generative}, domain adaptation~\cite{wang2018racial}, differences in face-based attributes across demographics~\cite{wang2018they}, and even data exploration~\cite{muthukumar2019}. Yin~\etal proposed to augment the feature space of underrepresented classes using other classes with a diverse collection of samples to encourage distributions of underrepresented classes to more closely resemble that of the others~\cite{yin2019feature}. Similarly, others formulated the imbalanced class problem as one-shot learning, where a \gls{gan} was trained to generate face features to augment classes with fewer samples~\cite{ding2018one}. \gls{gapf} was proposed to create fair representations of the data in a quantifiable way, allowing for the finding of a de-correlation scheme from the data without access to its statistics~\cite{huang2018generative}. Wang~\etal defined subgroups at a finer-level (\ie Chinese, Japanese, Korean), and determined the familiarity of faces inter-subgroup~\cite{wang2018they}. Genders have also been used to make subgroups (\eg for analysis of gender-based face encodings~\cite{muthukumar2019}). Most recently,~\cite{wang2018racial} proposed adapting domains to bridge the bias gap by knowledge transfer, which was supported by a novel data collection, \gls{rfw}. The release of \gls{rfw} occurred after \gls{bfw} was built - although similar in terms of demographics, \gls{rfw} uses faces from MSCeleb~\cite{guo2016ms} for testing, and assumes CASIA-Face~\cite{yi2014learning} and VGG2~\cite{Cao18} were used to train. In contrast, our \gls{bfw} assumes VGG2 as the test set. 
        % As shown in \cite{wang2018racial}, MSCeleb is highly imbalanced, primarily consisting of images of white individuals. 
        % For this, we expect bias from data, for this is the training set. 
        Furthermore, \gls{bfw} balances gender and race: \gls{rfw} splits subgroups by gender and race, while \gls{bfw} has gender, race, or both). 
        % Thus, \gls{rfw} and \gls{bfw} complement one another. % - both add demographic for subjects with faces in renowned, large-scale \gls{fr} datasets.

Most similar to us is~\cite{das2018, demogPairs, lopez2019dataset, srinivas2019face} - each were motivated by insufficient paired data for studying bias in \gls{fr}; then, problems were addressed using labeled data from existing image collections. Specifically, Hupont~\etal curated a set of faces based on racial demographics (\ie \gls{a}, \gls{b}, and \gls{w}) called \gls{dp}~\cite{demogPairs}, while~\cite{srinivas2019face} honed in on adults versus children called \gls{itwcc}. Like the proposed \gls{bfw}, both were built by sampling existing databases, but with the addition of tags for the respective subgroups of interest. Besides, the additional data of \gls{bfw} (\ie added an additional subgroup \gls{i}, along with additional subjects with more faces for all subgroups), we also further split subgroups by gender. Furthermore, we focus on the problem of facial verification and the different levels of sensitivity in cosine similarity scores per subgroup.

@@

            \multicolumn{2}{c}{Database} & \multicolumn{3}{c}{Number of}& \multicolumn{3}{c}{Balanced Labels}\\
            \cmidrule(lr){1-2}	\cmidrule(lr){3-5} \cmidrule(lr){6-8}
            Name & Source Data & Faces &  IDs & Subgroups & ID & Ethnicity & Gender\\\midrule
            \gls{dp}~\cite{demogPairs}     & CASIA-W~\cite{yi2014learning}, VGG~\cite{schroff2015facenet} \&VGG2~\cite{Cao18} & 10,800& 600 & 6 &\checkc& \checkc &\checkc \\
            \gls{rfw}~\cite{wang2018racial}     &  MS-Celeb-1M &$\approx$80,000&$\approx$12,000& 4 & \xmark & \checkc &\xmark \\
            \gls{bfw} (ours) & VGG2 & 20,000 & 800 &8 & \checkc & \checkc &\checkc \\\bottomrule
        \end{tabular}
        \label{tab:compared}
            \vspace{-5mm}
    \end{table*}

>>>>

\section{Background Information}
    \subsection{Bias in \gls{ml}}
        The progress and commercial value of \gls{ml} is exciting; however, the trust of society in \gls{ml} will take time to build due to inherent biases. The exact definitions and implications of bias vary between sources, as do its sources and types. A common theme is that bias hinders performance ratings in ways that skew to a particular sub-population-- the source varies, whether from humans~\cite{windmann1998subconscious}, data and label types~\cite{tommasi2017deeper}, \gls{ml} models~\cite{amini2019uncovering, kim2019learning}, or evaluation protocols~\cite{stock2018convnets}. For instance, a vehicle-detection model might miss cars if training data was mostly trucks. In practice, many \gls{ml} systems learn on biased data, which could be detrimental for society.

    \subsection{Bias in \gls{fr}}
        Problems of bias in \gls{fr} have been driven by different motivations; therefore, solutions have been posed for different matters. Instances can be in problems of data augmentation~\cite{yin2019feature}, one-shot learning~\cite{ding2018one}, demographic parity and fairness with priority on privacy~\cite{huang2018generative}, domain adaptation~\cite{wang2018racial}, differences in face-based attributes across demographics~\cite{wang2018they}, and even data exploration~\cite{muthukumar2019}. Yin~\etal proposed to augment the feature space of underrepresented classes using other classes with a diverse collection of samples to encourage distributions of underrepresented classes to more closely resemble that of the others~\cite{yin2019feature}. Similarly, others formulated the imbalanced class problem as one-shot learning, where a \gls{gan} was trained to generate face features to augment classes with fewer samples~\cite{ding2018one}. \gls{gapf} was proposed to create fair representations of the data in a quantifiable way, allowing for the finding of a de-correlation scheme from the data without access to its statistics~\cite{huang2018generative}. Wang~\etal defined subgroups at a finer-level (\ie Chinese, Japanese, Korean), and determined the familiarity of faces inter-subgroup~\cite{wang2018they}. Genders have also been used to make subgroups (\eg for analysis of gender-based face encodings~\cite{muthukumar2019}). Most recently,~\cite{wang2018racial} proposed to adapt domains to bridge the bias gap by knowledge transfer, which was supported by a novel data collection, \gls{rfw}. The release of \gls{rfw} occurred after \gls{bfw} was built - although similar in terms of demographics, \gls{rfw} uses faces from MSCeleb~\cite{guo2016ms} for testing, and assumes CASIA-Face~\cite{yi2014learning} and VGG2~\cite{Cao18} were used to train. In contrast, our \gls{bfw} assumes VGG2 as the test set. 
        % As shown in \cite{wang2018racial}, MSCeleb is highly imbalanced, primarily consisting of images of white individuals. 
        % For this, we expect bias from data, for this is the training set. 
        Furthermore, \gls{bfw} balances gender and race: \gls{rfw} splits subgroups by gender and race, while \gls{bfw} has gender, race, or both). 
        % Thus, \gls{rfw} and \gls{bfw} complement one another. % - both add demographic for subjects with faces in renowned, large-scale \gls{fr} datasets.

Most similar to us is~\cite{das2018, demogPairs, lopez2019dataset, srinivas2019face} - each were motivated by insufficient paired data for studying bias in \gls{fr}. Then, problems were addressed using labeled data from existing image collections. Specifically, Hupont~\etal curated a set of faces based on racial demographics (\ie \gls{a}, \gls{b}, and \gls{w}) called \gls{dp}~\cite{demogPairs}, while~\cite{srinivas2019face} honed in on adults versus children called \gls{itwcc}. Like the proposed \gls{bfw}, both were built by sampling existing databases, but with the addition of tags for the respective subgroups of interest. Aside from the additional data of \gls{bfw} (\ie added an additional subgroup \gls{i}, along with additional subjects with more faces for all subgroups), we also further split subgroups by gender. Furthermore, we focus on the problem of facial verification and the different levels of sensitivity in cosine similarity scores per subgroup.

@@

            \multicolumn{2}{c}{Database} & \multicolumn{3}{c}{Number of}& \multicolumn{3}{c}{Balanced Labels}\\
            \cmidrule(lr){1-2}	\cmidrule(lr){3-5} \cmidrule(lr){6-8}
            Name & Source Data & Faces &  IDs & Subgroups & ID & Ethnicity & Gender\\\midrule
            \gls{dp}~\cite{demogPairs}     & CASIA-W~\cite{yi2014learning}, VGG~\cite{schroff2015facenet} \&VGG2~\cite{Cao18} & 10,800& 600 & 6 &\checkc& \checkc &\checkc \\
            \gls{rfw}~\cite{wang2018racial}     &  MS-Celeb-1M &$\approx$80,000&$\approx$12,000& 4 & \xmark & \checkc &\xmark \\
            \gls{bfw} (ours) & VGG2 & 20,000 & 800 &8 & \checkc & \checkc &\checkc \\\bottomrule
        \end{tabular}
        \label{tab:compared}
%            \vspace{-5mm}
    \end{table*}
